---
title: "1. GPS-GSM data pre-filtering"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: true
    smooth_scroll: true
---

```{r setup, eval=TRUE, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Filtering of flight locations, speeds, and bursts (specific to GPS-GSM tracked birds)

# Libraries

```{r}

if (!require(readr)) install.packages("readr")
if (!require(readxl)) install.packages("readxl")
if (!require(dplyr)) install.packages("dplyr")
if (!require(lubridate)) install.packages("lubridate")
if (!require(raster)) install.packages("raster")
if (!require(SDLfilter)) install.packages("SDLfilter")
if (!require(sf)) install.packages("sf")

library(readr)
library(readxl)
library(dplyr)
library(lubridate)
library(raster)
library(SDLfilter)
library(sf)

```

# Set directory and folders

Set the working directory where the files are located

```{r}

setwd("C:/Users/...")
folder <- "C:/Users/..."

# Obtain the list of CSV and Excel files in the folder.
# The following lines open the CSV and Excel files in the folder
# Preferably, these should all be CSV files downloaded directly from Movebank (or Excel files if not available)
# If there are both CSV and Excel files, errors may occur when merging the databases

csv_files <- list.files(folder, pattern = "\\.csv$", full.names = TRUE)

```

# Functions

## Function to read CSV files

```{r}

read_file <- function(file) {
  
  if (grepl("\\.csv$", file)) {
    data <- read_csv(file, show_col_types = FALSE)
    }
  
  # Convert all columns to character type to avoid errors when merging
  data <- mutate_all(data, as.character)
  
  return(data)
  
  }

```

## Function to identify and filter the Bursts

Within a burst, keep data with \> satellite count and \< hdop

```{r}

process_burst <- function(df) {
  
  df$DateTime <- as_datetime(df$DateTime)   
  
  df$DateTime2 <- c(df$DateTime[1], df$DateTime[-length(df$DateTime)])
  
  df$diff <- difftime(df$DateTime, df$DateTime2)
  df$diff <- as.numeric(df$diff)
  
  condition <- df$diff > 3
  df$burst_ID <- cumsum(condition)
  
  df <- df %>%
    group_by(burst_ID) %>%
    filter(
      (is.na(`gps:satellite-count`) | `gps:satellite-count` == max(`gps:satellite-count`, na.rm = TRUE)) &
        (is.na(`gps:hdop`) | `gps:hdop` == min(`gps:hdop`, na.rm = TRUE))
    ) %>%
    ungroup()
  
  df <- df %>%
    distinct(burst_ID, .keep_all = TRUE) %>%
    dplyr::select(-DateTime2)
  
  return(df)
  
  }

```

# Load and prepare data

```{r}

# Load necessary files

elevation <- raster("C:/Users/.../elevation.tif")

data_list <- lapply(c(csv_files), read_file) 
unified_data <- bind_rows(data_list)

# Convert variables

unified_data$species_name <- as.factor(unified_data$`individual-taxon-canonical-name`) 

unified_data$id <- as.factor(unified_data$`individual-local-identifier`) 

unified_data$DateTime <- as.POSIXct(unified_data$timestamp, format = "%Y-%m-%d %H:%M:%S", tz = "UTC") 

unified_data$lon <- as.numeric(unified_data$`location-long`) 
unified_data$lat <- as.numeric(unified_data$`location-lat`) 

unified_data$`height-above-ellipsoid` <- as.numeric(unified_data$`height-above-ellipsoid`) 
unified_data$`height-above-msl` <- as.numeric(unified_data$`height-above-msl`) 
unified_data$`height-raw` <- as.numeric(unified_data$`height-raw`) 

unified_data <- unified_data %>% mutate(line_number = row_number())

```

# Remove outliers

Remove points with latitude and longitude equal to 200 (Movebank outliers) Remove rows where DateTime is NA

```{r}

unified_data <- unified_data %>%
  dplyr::filter(!(lat == 200 & lon ==200)) %>%
  dplyr::filter(!(is.na(DateTime))) %>%
  arrange(id, DateTime)

```

# Loop for each species

```{r}

# List to store results

results <- list()

for (individual in unique(unified_data$`id`)) {
  
  individual_data <- subset(unified_data, `id` == individual)
  
  # Species-specific: Tetrax tetrax Apply different code according to the species
  
  if (unique(individual_data$species_name) == "Tetrax tetrax") {
    
    # 1. Filter the Bursts
    individual_data <- process_burst(individual_data)
    
    # 2. Filter of heights
    # In cases where there is no height above sea level -> calculate it using the Digital Elevation Model
    
    # Extract the 'elevation' value for each location
    individual_data$elevation_value <- extract(elevation,
                                               cbind(individual_data$'lon',
                                                     individual_data$'lat'))
    
    # Fill 'height-above-msl' where it is NA
    
    individual_data$`height-above-msl` <- ifelse(
      is.na(individual_data$`height-above-msl`), 
      individual_data$`height-above-ellipsoid` - individual_data$elevation_value,
      individual_data$`height-above-msl`
      )
    
    # Filter data where height-above-msl is within the range of elevation ±20 m
    # (species-specific: adjust the desired range for other species)
    
    individual_data <- individual_data %>%
    filter(`height-above-msl` >= elevation_value - 20 & `height-above-msl` <= elevation_value + 20)
      
    # 3. Filter locations with repeated DateTime and coordinates
      
    individual_data <- individual_data[order(individual_data$DateTime),]
      
    duplicatedLocs <- which(individual_data$DateTime[1:(nrow(individual_data)-1)] == individual_data$DateTime[2:(nrow(individual_data))])
      
    if (length(duplicatedLocs) >= 1) {individual_data <- individual_data[-duplicatedLocs,] }
      
    } 
  
  else if (unique(individual_data$species_name) == "Burhinus oedicnemus") {
    
    # (Repeat similar steps for each species)
    # Code for "Burhinus oedicnemus"
    
    }
  
  # Save in the result list
  
  results[[individual]] <- individual_data
  
  }


# Combine all processed data into a single dataframe

processed_data <- do.call(rbind, results)

```

# Review results

```{r}

head(processed_data)

table(unified_data$species_name)
table(processed_data$species_name)

```

# Filter by years 2022 and 2023 and write csv

```{r}

processed_data <- processed_data %>% mutate(timestamp = as.POSIXct(timestamp))

processed_data <- processed_data %>% filter(format(timestamp, "%Y") %in% c("2022", "2023"))

write.csv(processed_data, "preprocessed_data.csv", row.names = FALSE)

```
